Retrofitting Effect Handlers onto OCaml

### 2021-03-12 retro-concurrency.tex
updates

\newcommand{\farg}[2]{\langle #1 ~#2 \rangle_a}
\newcommand{\ffun}[1]{\langle #1 \rangle_f}
\newcommand{\faritha}[3]{\langle #1 ~#2 ~#3 \rangle_{b1}}
\newcommand{\farithb}[2]{\langle #1 ~#2 \rangle_{b2}}

@@

\newcommand{\cstack}{\gamma} % C Stack
\newcommand{\ostack}{\omega} % OCaml Stack
\newcommand{\cstacka}[2]{\big \lceil #1, #2 \big \rceil_c} % C stack with arguments
\newcommand{\ostacka}[2]{\big \lceil #1, #2 \big \rceil_o} % OCaml stack with arguments
\newcommand{\ostackemp}{\bullet}
\newcommand{\stack}{\sigma}

@@

\newcommand{\ostep}{\xrightarrow{o}}
\newcommand{\cstep}{\xrightarrow{c}}
\newcommand{\step}{\rightarrow}
\newcommand{\multistep}{\rightarrow\!\!*}

@@

\author{KC Sivaramakrishnan}
\affiliation{
  \institution{IIT Madras}            %% \institution is required
	\country{India}
}
\email{kcsrk@cse.iitm.ac.in}          %% \email is recommended

\author{Stephen Dolan}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{stephen.dolan@cl.cam.ac.uk}

\author{Leo White}
\affiliation{
  \institution{Jane Street}            %% \institution is required
	\country{UK}
}
\email{leo@lpw25.net}

\author{Sadiq Jaffer}
\affiliation{
  \institution{Opsian and OCaml Labs} %% \institution is required
	\country{UK}
}
\email{sadiq@toao.com}

\author{Tom Kelly}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{tom.kelly@cantab.net}

\author{Anil Madhavapeddy}
\affiliation{
  \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
	\country{UK}
}
\email{avsm2@cl.cam.ac.uk}          %% \email is recommended

@@

\subsection{Asynchronous I/O}

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. Let us illustrate the power of
effect handlers in OCaml by implementing a simple lightweight thread library
with support for forking new threads and yielding control to other threads. We
will then extend this scheduler with support for synchronously reading from
channels and subsequently make it asynchronous without changing the client code
for asynchrony.

Many programming languages, including OCaml, require non-local change to the
code in order to support asynchronous I/O, often leading to a dichotomy
synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
primitive support for concurrency, users will have to use monadic I/O libraries
such as Lwt~\cite{lwt} and Async~\cite{async}. Languages such as GHC Haskell
and Go provide lightweight threads but bake their implementation into the
runtime system. With effect handlers, asynchronous I/O can be implemented
directly in OCaml.

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. Our continuations are
assumed to be resumed exactly once using |continue| or |discontinue|. As a
result, when a computation performs an effect, we are guaranteed that the
control will return. For the non-returning cases (value and exceptional
return), the code already handles resource cleanup. OCaml does not have a
\texttt{\footnotesize try/finally} construct commonly used for resource cleanup
in many programming languages. Alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are implemented using exceptions.
Thus, the linear use of continuations enabled by the |discontinue| primitive
ensures backwards compatibility of legacy OCaml systems code under non-local
control flow introduced by effect handlers. Leijen~\cite{Leijen18} explicitly
extends effect handlers with |initially| and |finally| clauses in Koka for
resource safety. Dolan et al.~\cite{TFP17} describe the interaction
of effect handlers and asynchronous exceptions. This is orthogonal to the
contributions of this paper. Our focus is the compiler and runtime system
support for implementing effect handlers.

>>>>

\newcommand{\farg}[2]{\langle #1 ~#2 \rangle_{\text a}}
\newcommand{\ffun}[1]{\langle #1 \rangle_{\text f}}
\newcommand{\faritha}[3]{\langle #1 ~#2 ~#3 \rangle_{\text b1}}
\newcommand{\farithb}[2]{\langle #1 ~#2 \rangle_{\text b2}}

@@

\newcommand{\cstack}{\gamma} % C Stack
\newcommand{\ostack}{\omega} % OCaml Stack
\newcommand{\cstacka}[2]{\big \lceil #1, #2 \big \rceil_{\text c}} % C stack with arguments
\newcommand{\ostacka}[2]{\big \lceil #1, #2 \big \rceil_{\text o}} % OCaml stack with arguments
\newcommand{\ostackemp}{\bullet}
\newcommand{\stack}{\sigma}

@@

\newcommand{\ostep}{\xrightarrow{\text o}}
\newcommand{\cstep}{\xrightarrow{\text c}}
\newcommand{\step}{\rightarrow}
\newcommand{\multistep}{\rightarrow\!\!*}

@@

\author{KC Sivaramakrishnan}
\affiliation{
  \institution{IIT Madras}            %% \institution is required
	\city{Chennai}
	\country{India}
}
\email{kcsrk@cse.iitm.ac.in}          %% \email is recommended

\author{Stephen Dolan}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\city{Cambridge}
	\country{UK}
}
\email{stephen.dolan@cl.cam.ac.uk}

\author{Leo White}
\affiliation{
  \institution{Jane Street}            %% \institution is required
	\city{London}
	\country{UK}
}
\email{leo@lpw25.net}

\author{Sadiq Jaffer}
\affiliation{
  \institution{Opsian and OCaml Labs} %% \institution is required
	\city{Cambridge}
	\country{UK}
}
\email{sadiq@toao.com}

\author{Tom Kelly}
\affiliation{
  \institution{OCaml Labs}            %% \institution is required
	\city{Cambridge}
	\country{UK}
}
\email{tom.kelly@cantab.net}

\author{Anil Madhavapeddy}
\affiliation{
  \institution{University of Cambridge Computer Laboratory and OCaml Labs}            %% \institution is required
	\city{Cambridge}
	\country{UK}
}
\email{avsm2@cl.cam.ac.uk}          %% \email is recommended

@@

\subsection{Asynchronous I/O}
\label{sec:aio}

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to support concurrent programming. To
illustrate this, let us implement a simple lightweight thread library with
support for forking new threads and yielding control to other threads. We will
then extend this scheduler with support for synchronously reading from channels
and subsequently make it asynchronous without changing the client code for
asynchrony.

Many programming languages, including OCaml, require non-local change to the
code in order to support asynchronous I/O, often leading to a dichotomy
synchronous and asynchronous code~\cite{colour}. Since OCaml does not have
primitive support for concurrency, users will have to use monadic I/O libraries
such as Lwt~\cite{lwt} and Async~\cite{async}, where all asynchronous code must
use monad syntax. Languages such as GHC Haskell and Go provide lightweight
threads, which avoids the dichotomy between synchronous and asynchronous code.
However, these languages bake-in the lightweight thread implementation into the
runtime system. With effect handlers, asynchronous I/O can be implemented
directly in OCaml as a library.

@@

Multicore OCaml builds on the existing defensive coding practices against
exceptions to clean up resources on non-returning exits. We assume that the
continuations are resumed exactly once using |continue| or |discontinue|. Under
this assumption, when a computation performs an effect, we expect the control
to return. For the non-returning cases (value and exceptional return), the code
already handles resource cleanup.

OCaml does not have a \texttt{\footnotesize try/finally} construct commonly
used for resource cleanup in many programming languages. The OCaml standard
library~\cite{FunProtect} as well as alternative standard libraries such as
Base~\cite{BaseProtect} and Core~\cite{CoreProtect} provide mechanisms
analogous to |unwind-protect|, which are in turn implemented using exception
handlers. Thus, the linear use of continuations enabled by the |discontinue|
primitive ensures backwards compatibility of legacy OCaml systems code under
non-local control flow introduced by effect handlers.

Leijen~\cite{Leijen18} explicitly extends effect handlers with |initially| and
|finally| clauses in Koka for resource safety. Dolan et al.~\cite{TFP17}
describe the interaction of effect handlers and asynchronous exceptions. This
is orthogonal to the contributions of this paper. Our focus is the compiler and
runtime system support for implementing effect handlers.
